Exploiting delayed dichromatism to disentangle the effects of adult survival and recruitment on the population dynamics of waterfowl

### 2023-02-09 Code/Draft_ruddy_duck_vital_rates.Rmd
update of all figures, ready for results

Assessing population growth-rate is a key step towards a better understanding of factors underlying the dynamics of natural populations [e.g. @Niel2005]. It is also crucial for measuring the efficacy of management actions eventually undertaken to help population recovery, control, or eradication [@Shea1998]. Among the approaches available to managers for reaching these goals, those relying on the monitoring of the number of individuals are by far the most popular [@Rintala2022]. In many instances, these methods even enable environmental factors underlying changes in population size to be investigated and hence some mitigating actions to be implemented [@Faillettaz2019]. Unfortunately, relying on the monitoring of numbers of individuals alone hinder basic understanding of the demographic mechanisms underlying changes in population growth rates. More specifically, based on counts only, it is nearly impossible to assess to what extent population growth rate is affected by change in survival rather than to variations in reproductive parameters [@Austin2000]. To assess the relative sensitivity of population growth rate to factors affecting survival or reproductive parameters, demographers most often rely on the monitoring of individuals by capture-mark-recapture [@Lebreton1992]. In practice however, one cannot always rely on the capture and release of individuals owing to the legal status of the species that precludes any additional disturbance, e.g. critically endangered species, or forbid the release of alive individuals, e.g. invasive species or pests. Indeed, although highly efficient for assessing demographic parameters, CMR methods have potential drawbacks such as being invasive and hardly affordable when time and money are limiting. Genetic monitoring is a non-invasive alternative to CMR approaches based on physical captures, but it suffers from being costly and time-consuming and requires quite rigorous sampling schemes in the field.

@@

The ruddy duck is a stiff-tailed duck native from the Americas. From seven individuals initially acclimated in the Slimbridge Wetland Centre in the Great Britain in the 40’s [@GutierrezExposito2020], a feral population began to establish with the first observed reproductive attempts in the wild in the 60s (Figure \@ref(fig:count)). This feral population rapidly grew and spread to the entire country to reach more than 5000 individuals in the early 2000s. By the end of the 80s, a feral population began to establish also on the continent in France, supposedly owing to the arrival of individuals born in Great Britain. Contrary to what was observed in Great Britain, the French Ruddy duck population did not spread much, with the vast majority of observations and breeding attempts clustering in Western France. In France, during winter, almost none Ruddy duck is observed outside the lake of Grand Lieu (47.09°N, 1.67°W), which is thus the exclusive wintering site and greatly facilitates the monitoring of this population. The feral European populations of Ruddy duck are considered as a major threat to the endangered native White-headed duck population of the South-Western Mediterranean because of hybridization risks and thus an elevated potential for genetic pollution and eventually genetic assimilation of the latter by the former [@MunozFuentes2007]. In order to mitigate the risks of genetic pollution of the White-headed duck by Ruddy ducks, a European plan of eradication of Ruddy ducks has been adopted by the European Commission. As a result, eradication measures were taken both in Great Britain and in France from the late 90s [@GutierrezExposito2020], (Figure \@ref(fig:count)). The release of non-native species in the wild, including Ruddy ducks, is forbidden in countries of the European Union, and the status of White-headed duck populations is highly unfavourable. The use of capture-mark-recapture to monitor populations of these species is thus impossible. As a result, the effectiveness of the eradication of Ruddy ducks and management actions intended to favour the recovery of White-headed ducks were essentially checked through censuses. Censuses, however, prevented from assessing the relative effects or variations in survival and reproductive success/recruitment on population growth rate, which is a key towards a proper understanding of the efficiency of management actions.

```{r count, fig.cap = "(ref:count)"}
knitr::include_graphics("../Output/plot_pres_2.png")
```

Like many other ducks, Ruddy ducks and White-headed ducks both display delayed dichromatism. Thereby immature males acquire the typical colourful plumage of their kind over the course of the interbreeding season, which is typically in late winter. Immature males are thus confounded to immature and adult females before their moult (Figure \@ref(fig:flock)). The delayed dichromatism explains why the apparent proportion of females in the winter counts is always higher than the true proportion of females that can be observed in winter hunting bags (Figure \@ref(fig:pfem)). Thus, it is possible to assess the proportion of immatures within a population by monitoring the seasonal evolution of apparent sex ratio and assuming an even sex ratio at birth, comparable mortality rates between females and males among juveniles, and unchanging adult sex ratio over the monitoring period. Then, both adult survival rate and recruitment rate can be estimated by combining the age ratio thus obtained with reliable counts over two successive winters.

@@

```{r pfem, fig.cap = "(ref:pfem)"}
knitr::include_graphics("../Output/plot_pres_1.png")
```

We monitored the demography of Ruddy duck populations in both Great-Britain and France thanks to exhaustive counts on the wintering grounds during the period between December 1 and January 31. In addition to these exhaustive counts, censuses that distinguished female-like from male-like individuals were performed on the same period from 2007 to 2012 in Great Britain, which corresponded to a strong decline of the population. In France, the monitoring spread over 1999, 2001-2009, and 2012-2019, which corresponded to a stabilized population size excepted over 2001-2004, for which the population grew rapidly without control effort. During these years, the eradication effort, under control of the authorities in both countries, was also scrupulously recorded.

@@

In waterfowl, the survival is commonly higher in adult males than in adult females. Adult sex ratio is then generally biased towards males (Figure \@ref(fig:pfem), @Wood2021). As a consequence, deducing the immature proportion from the observed proportion of adult males is not straightforward because the adult proportion is not just twice as much as the observed proportion of adult males. There was no counts differentiating male-like from female-like during the breeding period to estimate properly adult sex ratio. But sex identification in adults from hunting bags was available for both Ruddy duck populations. However, the small population size in France prevented from getting precise adult sex ratios. A comparison over months in UK showed no difference in male proportion, so data from adults collected over the whole year were used to estimate the male proportion. A comparison of male proportion among years with more than 500 samples did not show significant interannual differences. Even if the adult sex ratio may vary over a long time range, it is relatively stable over a few years [@Wood2021]. As a consequence, hunting bag data from all adults were pooled to estimate the male proportion in adults, see Equation \@ref(eq:pm).

@@

The relevance of the method was assessed by testing the likelihood of the estimates of vital rates. For the adult survival rate, we checked if the values belonged to the expected interval $[0;1]$, and we compared the average values to literature data, since the adult survival is commonly assessed by capture mark recapture [@Lebreton2001]. For the recruitment rate, which is rarely explicitly informed in the literature, we checked if the values were equal or below the maximum recruitment rate expected without exploitation and without negative density-dependence processes. If the recruitment rate outputs of the model are significantly superior to the maximum recruitment rate, it means that the method overestimates the recruitment rate, which reflects that a part of adult males are not detected during count surveys. A null or negative recruitment rate would mean that immature males moulted before the survey or/and the assumption on the stability of the adult sex ratio is overrated.

The maximum recruitment rate is deduced by using its relationship to the maximum growth rate and the maximum survival, see Equation \@ref(eq:GR). This relationship comes from a simple reasoning for a closed population: the population size in year $t$ is equal to the number of adults that survived over year $t - 1$ plus the offspring produced in year $t - 1$ that survived until the reproduction period of year $t$, i.e. the recruitment in year $t$. The growth rate of a population is thus the sum of the adult survival rate and the recruitment rate [@Flint2015]. For an open population, adult survival and recruitment rate are confounded to adult and recruit migration respectively. This relationship becomes more complex if a species with delayed maturity is considered, see @Robertson2008.

@@

\begin{align} 
N_{t} & = N_{t-1} - D_{t-1} + R_{t} \notag \\
N_{t} & = N_{t-1} - (1 - s_{t}).N_{t-1} + r_{t}.N_{t-1} \notag \\
N_{t} & = s_{t}.N_{t-1} + r_{t}.N_{t-1} \notag \\
\frac{N_{t}}{N_{t-1}} & = s_{t} + r_{t} \notag \\
\lambda_{t} & = s_{t} + r_{t} 
(\#eq:GR)
\end{align}

The maximum growth rate occurred for both populations during their expansion phase, when the evolution of the population size was the steepest. To estimate robust maximum population growth rates for both populations, we smoothed annual population growth rate over a consistent time period by using a linear regression on the logarithm scale, see Equation \@ref(eq:L).

@@

Practically, we used a very conservative prior on the maximum adult survival by setting a uniform distribution on $[0.7; 0.9]$. The lower limit corresponds to the upper range of survival rates observed in literature for species of similar weight, the upper limit corresponds to the range of survival rates of long-life species [@Nichols1997; @Krementz1997; @Buxton2004]. For the maximum growth rate, we rejected Great Britain data before 1972 since the size estimate of this population is noisy below 50 individuals (Figure \@ref(fig:count)). After reaching the threshold of 1000 individuals, the Great Britain population growth observed a strong inflexion whereas no harvest pressure was applied. This observation suggests that a negative density-dependence process might occur beyond 1000 individuals and led to consider only the first sequence of the time series to infer on the maximum growth rate in Great Britain. For the French population, the two sequences without harvest were used, i.e. 1994-1999 and 2001-2004.

@@

Harvest strategies differed from Great Britain to France. In Great Britain, harvest occurred in winter, i.e. before the breeding period, whereas it was done mostly in summer in France, i.e. during and after the breeding period. To identify the impact of the two strategies on the population dynamics, we assessed how vital rates responded to the harvest by considering both average drops and variability changes. To complete our understanding on why the Great Britain strategy was more performing on controlling Ruddy ducks, we compared vital rates with and without harvest in light of the underlying average population growth rates.

For all the sub-models in Section \@ref(MM), the Bayesian framework was used for its efficiency and simplicity to propagate error through the parameters. We used uninformative priors on all parameters. We generated three chains of length 500000, with a thinning of 10 to avoid autocorrelation in the samples, and we discarded the first 2000 samples as burn-in. Chain convergence was assessed using the Gelman and Rubin convergence diagnostic (R<1.1, @Gelman1992). We fit the models using NIMBLE [@Valpine2017] run from R [@RCT2022]. Data and code are available here: https://github.com/adri-tab/Ruddy_duck_vital_rates. The values **X[Y; Z]** reported in results are the medians and the associated boundaries of the 95% confidence interval of posterior distributions. The median was preferred to the mean because of its robustness to skewed distribution.

@@

@@

Time series derived from count data show that the two populations display comparable initial growth (Figure \@ref(fig:count2)). They grew freely in both countries until 1999. Although shooting was initiated in 1999 in both countries, both populations begun declining only after much higher harvest efforts had been achieved in 2005 and 2018 respectively (figure 3). This history underlines that both populations observed a large spectrum of harvest effort, from no pressure to very high pressure. For sake of interpretation, we ascribed the time-period 2001-2004 to the no harvest category in France because the harvest pressure was negligible hand hence had no effect on the population growth.

Also, covering contrasted populations dynamics corresponding to different levels of harvesting pressure (Figure 2.3). In France, data before 2004 correspond to a period of steady population growth with (almost) no harvest pressure, whereas data in the following years correspond to a stabilized population size with a significant level of harvest pressure (Figure 2.3). In UK, the counting data correspond to a quick population depletion associated to a high level of harvest pressure, especially before the reproduction period. In parallel, the sampling dataset, which corresponds to individuals shot in winter during control operations, covers 9 years only in UK, with 5 years in common with the corresponding counting time series.
Estimating recruitment rate from the evolution of apparent sex-ratio in winter

The counting dataset, which differentiates the male-like from the female-like individuals in winter, covers 7 years and 19 years in UK and France respectively (Figure \@ref(fig:count)). Combined together, the counting datasets cover different population dynamics corresponding to a gradient of harvest pressure. In France, data before 2004 correspond to a period of population growth with (almost) no harvest pressure, whereas data in the following years correspond to a stabilized population size with a significant level of harvest pressure (Figure \@ref(fig:count)). In UK, the counting data correspond to a quick population depletion associated to a high level of harvest pressure, especially before the reproduction period. In parallel, the sampling dataset, which corresponds to individuals shot in winter during control operations, covers 9 years only in UK, with 5 years in common with the corresponding counting time series.

## Recruitment rate variability

The male proportion in adults, which is necessary to infer on the recruitment rate from count data, is estimated at 0.60 [0.59; 0.61]. The counting method provides estimations of the recruitment rate, see Figure \@ref(fig:rrts)A. In UK, values vary from 0.06 [0.05; 0.07] to 0.27 [0.20; 0.34] recruits per breeder, and in France, 0.08 [0.04; 0.12] to 0.76 [0.64; 0.88] recruits per breeder. There are significant year-to-year variations with a maximum amplitude of 0.68 recruits per breeder in France (9.5-time variability). On the available period, recruitment rate in UK is smaller than in France. Recruitment rates are estimated in both populations over 5 years, and no correlation is observed between the two populations. The lowest recruitment rate observed in France, which is in 2019, corresponds to the start of the depletion of the population. It is noticeable that it is in the same range than the UK values, also estimated during a strong population depletion.

Recruitment rates estimated from UK sample data are consistently higher than estimates from the corresponding count data, Figure \@ref(fig:rrts)A. The correlation between the two methods is strong and points out that the sampling method overestimates the recruitment rate by 44% compared to the counting method, Figure \@ref(fig:rrts)B. Estimation precision is similar for the two methods. The UK population started to deplete from 2006, but no significant shift is observed between the two periods.

```{r rrts, fig.cap = "(ref:rrts)"}
knitr::include_graphics("../Output/plot_4.png")
```

## Consistency of the corresponding adult survival

Adult survival rates were deduced from the recruitment rate estimates and the observed population growth rates, see Figure \@ref(fig:srts)A. In UK, values that are obtained from the counting method ranged from 0.21 [0.20; 0.22] to 0.68 [0.65; 0.72]. In France, survival rates vary from 0.39 [0.34; 0.44] to 1.08 [0.98; 1.17]. Like recruitment rates, there is significant year-to-year variability without clear temporal structure in both populations with a maximum amplitude of 0.69 in France (2.8-time variability). Unlike the recruitment rate, the survival rate in 2019 in France is not a low outlier.

As expected, survival rates estimated from UK sample data are consistently lower than estimates from the corresponding count data, Figure \@ref(fig:srts)A. The correlation between the two methods points out that the sampling method underestimates the survival rate by 32% compared to the counting method, Figure \@ref(fig:srts)B. Estimation precision is similar for the two methods. The sample time series shows that the adult survival is significantly lower after 2006, which corresponds to the start of the depletion of the UK population.

```{r srts, fig.cap = "(ref:srts)"}
knitr::include_graphics("../Output/plot_5.png")
```

## Comparison to maximum recruitment rate

```{r mgr, fig.cap = "(ref:mgr)"}
knitr::include_graphics("../Output/plot_6.png")
```

Upon the conservative assumption of a maximum adult survival uniformly distributed over 0.70 to 0.90, the maximum recruitment rate reaches 0.65 [0.51; 0.75] and 0.66 [0.45; 0.90] recruits per breeder for UK and France respectively (Figure \@ref(fig:grcomp)). When there is no harvest, the estimated average recruitment rate of the French population is in the range of the proxy of the maximum recruitment rate, 0.51 [0.46; 0.57] recruits per breeder. When there is harvest, the average recruitment rates are 0.14 [0.12; 0.16] and 0.33 [0.31; 0.36] recruits per breeder for the UK and France respectively. Both values are significantly lower than the maximum recruitment rate, and it is noticeable that the average recruitment rate is lower for the UK population, which is in depletion over the count time series, than for the French population, which of the dynamic is mostly stable over the monitored period.

```{r grcomp, fig.cap = "(ref:grcomp)"}
knitr::include_graphics("../Output/plot_7.png")
```

@@

Assuming that the sample data provide a good picture of the interannual variability of the recruitment rate, the good correlation with the outputs of the counting method demonstrates the ability of this method to track the recruitment rate variability. This result is quite robust because the correlation is obvious with only a five-year time series. The assumption of a constant adult sex ratio over time is then acceptable because a significant change in adult sex ratio among years would have biased the evolution of the recruitment rate. Since this ratio integrates many age-cohorts, the temporal autocorrelation is structurally strong. Significant variations can then occur only every multiple years, especially if the species is long-living. Consequently, it is not necessary to monitor and update the adult sex ratio on a yearly base. The adult sex ratio found here on a restricted time period is in line with the ruddy duck in its native area [0.62 in @Bellrose1980] and with other duck species [@Wood2021]. The robustness of the correlation between the two methods might be overrated because two successive population size values are parts of the recruitment rate, which mechanically scale the outputs. It explains why the correlation on the immature proportion, the other component of the recruitment rate, is a bit more noisy than for the recruitment rate (Figure \@ref(fig:sup)A). The uncertainty around the estimates is similar for the two methods even though the counting method is more indirect to infer on the recruitment rate, which is satisfying.

@@

# Supplement

```{r sup, fig.cap = "(ref:sup)"}
knitr::include_graphics("../Output/plot_8.png")
```

(ref:count) Evolution of the size of the two Ruddy duck populations; y-axis is on a log-scale

@@

(ref:pfem) Female proportion in hunting bags vs in counts; For the first category, one point corresponds to the female proportion estimated from direct sex identification of the Ruddy ducks controlled over one year in a population; for the second category, one point is the proportion of female-like plumage in a population counted in winter. The difference is due to immature males that look like females before moulting

(ref:grts) Evolution of the size of the two Ruddy duck populations in light of the harvest pressure; on time periods without harvest, maximum population growth rates can be estimated (slopes in red for the UK and in blue for France)

(ref:rrts) (A) Estimates of recruitment rate, i.e. average number of recruits produced per breeder; (B) Comparison of the recruitment rate values between the two estimation methods; bars define the 95% confidence intervals

(ref:srts) (A) Estimates of adult survival rate using population growth rates; (B) Comparison of the survival rate values between the two estimation methods; bars define the 95% confidence intervals

>>>>

Assessing population growth-rate is a key step towards a better understanding of factors underlying the dynamics of natural populations [e.g. @Niel2005]. It is also crucial for measuring the efficacy of management actions eventually undertaken to help population recovery, control, or eradication [@Shea1998]. Among the approaches available to managers for reaching these goals, those relying on the monitoring of the number of individuals are by far the most popular [@Rintala2022]. In many instances, these methods even enable environmental factors underlying changes in population size to be investigated and hence some mitigating actions to be implemented [@Faillettaz2019]. Unfortunately, relying on the monitoring of numbers of individuals alone hinder basic understanding of the demographic mechanisms underlying changes in population growth rates. More specifically, based on counts only, it is nearly impossible to assess to what extent population growth rate is affected by change in survival rather than to variations in reproductive parameters [@Austin2000]. To assess the relative sensitivity of population growth rate to factors affecting survival or reproductive parameters, demographers most often rely on the monitoring of individuals by capture-mark-recapture [@Lebreton1992]. In practice however, one cannot always rely on the capture and release of individuals owing to the legal status of the species that precludes any additional disturbance, e.g. critically endangered species, or forbid the release of alive individuals, e.g. invasive species or pests. Indeed, although highly efficient for assessing demographic parameters, capture-mark-recapture methods have potential drawbacks such as being invasive and hardly affordable when time and money are limiting. Genetic monitoring is a non-invasive alternative to capture-mark-recapture approaches based on physical captures, but it suffers from being costly and time-consuming and requires quite rigorous sampling schemes in the field.

@@

The Ruddy duck is a stiff-tailed duck native from the Americas. From seven individuals initially acclimated in the Slimbridge Wetland Centre in the Great Britain in the 40’s [@GutierrezExposito2020], a feral population began to establish with the first observed reproductive attempts in the wild in the 60s (Figure \@ref(fig:count)). This feral population rapidly grew and spread to the entire country to reach more than 5000 individuals in the early 2000s. By the end of the 80s, a feral population began to establish also on the continent in France, supposedly owing to the arrival of individuals born in Great Britain. Contrary to what was observed in Great Britain, the French Ruddy duck population did not spread much, with the vast majority of observations and breeding attempts clustering in Western France. In France, during winter, almost none Ruddy duck is observed outside the lake of Grand Lieu (47.09°N, 1.67°W), which is thus the exclusive wintering site and greatly facilitates the monitoring of this population. The feral European populations of Ruddy duck are considered as a major threat to the endangered native White-headed duck population of the South-Western Mediterranean because of hybridization risks and thus an elevated potential for genetic pollution and eventually genetic assimilation of the latter by the former [@MunozFuentes2007]. In order to mitigate the risks of genetic pollution of the White-headed duck by Ruddy ducks, a European plan of eradication of Ruddy ducks has been adopted by the European Commission. As a result, eradication measures were taken both in Great Britain and in France from the late 90s [@GutierrezExposito2020], (Figure \@ref(fig:count)). The release of non-native species in the wild, including Ruddy ducks, is forbidden in countries of the European Union, and the status of White-headed duck populations is highly unfavourable. The use of capture-mark-recapture to monitor populations of these species is thus impossible. As a result, the effectiveness of the eradication of Ruddy ducks and management actions intended to favour the recovery of White-headed ducks were essentially checked through censuses. Censuses, however, prevented from assessing the relative effects or variations in survival and reproductive success/recruitment on population growth rate, which is a key towards a proper understanding of the efficiency of management actions.

```{r count, fig.cap = "(ref:count)"}
knitr::include_graphics("../Output/plot_paper_2.png")
```

Like many other ducks, Ruddy ducks and White-headed ducks both display delayed dichromatism. Thereby immature males acquire the typical colourful plumage of their kind over the course of the interbreeding season, which is typically in late winter. Immature males are thus confounded to immature and adult females before their moult (Figure \@ref(fig:flock)). The delayed dichromatism explains why the apparent proportion of females in the winter counts is always higher than the true proportion of females that can be assessed in winter hunting bags (Figure \@ref(fig:pfem)). Thus, it is possible to assess the proportion of immatures within a population by monitoring the seasonal evolution of apparent sex ratio and assuming an even sex ratio at birth, comparable mortality rates between females and males among juveniles, and unchanging adult sex ratio over the monitoring period. Then, both adult survival rate and recruitment rate can be estimated by combining the age ratio thus obtained with reliable counts over two successive winters.

@@

```{r pfem, fig.cap = "(ref:pfem)"}
knitr::include_graphics("../Output/plot_paper_1.png")
```

We monitored the demography of Ruddy duck populations in both Great-Britain and France thanks to exhaustive counts on the wintering grounds during the period between December 1 and January 31. In addition to these exhaustive counts, censuses that distinguished female-like from male-like individuals (Figure \@ref(fig:count)) were performed from 2006 to 2012 in Great Britain, which corresponded to a strong decline of the population. In France, the monitoring spread over 1999, 2001-2009, and 2012-2019. This population grew rapidly during the first years of the monitoring and then stabilized from 2006, which corresponds to a period with a strong control effort. During these years, the eradication effort, under control of the authorities in both countries, was recorded. Even if the age ratio in the hunting bag was unknown before 2006 in France, we can qualify the harvest pressure on adults in three categories for both countries: "no harvest" before 1999, "low harvest" between 1999 and 2005, and "high harvest" from 2006 (see Figure \@ref(fig:sup) in Supplementary Materials).

@@

In waterfowl, the survival is commonly higher in adult males than in adult females. Adult sex ratio is then generally biased towards males (Figure \@ref(fig:pfem), @Wood2021). As a consequence, deducing the immature proportion within a population from the observed proportion of adult males is not straightforward because the adult proportion is not just twice as much as the observed proportion of adult males. There were no counts differentiating male-like from female-like during the breeding period to estimate properly adult sex ratio. But sex identification in adults from hunting bags was available for both Ruddy duck populations. However, the small population size in France prevented from getting precise adult sex ratios. A comparison over months in UK showed no difference in adult male proportion, so data from adults collected over the whole year were used to estimate the male proportion. A comparison of male proportion among years with more than 500 samples did not show significant interannual differences. Even if the adult sex ratio may vary over a long time range, it is relatively stable over a few years [@Wood2021]. As a consequence, hunting bag data from all adults were pooled to estimate the male proportion in adults, see Equation \@ref(eq:pm).

@@

The relevance of the method was assessed by testing the likelihood of the estimates of both vital rates. For the adult survival rate, we checked if the values belonged to the expected interval $[0;1]$, and we compared the values to literature data that are accessible because the adult survival is commonly assessed by capture-mark-recapture [@Lebreton2001]. The maximum adult survival is defined as a uniform distribution on $[0.7; 1]$. The lower limit corresponds to the upper range of survival rates observed in literature for waterfowl species of similar weight, the upper limit was set to 1 because the survival rates of long-life waterfowl species are very high [@Nichols1997; @Krementz1997; @Buxton2004]. The recruitment rate is more tricky to validate because it is not upper bounded as it is defined on $[0;\infty[$ and is neither explicitly informed in the literature. We then developed an approach to estimate the maximum recruitment rate expected without exploitation and without negative density-dependence processes and we checked if all values were equal or below this maximum recruitment rate. If the recruitment rate outputs of the model are significantly higher than the maximum recruitment rate, it means that the method overestimates the recruitment rate, which reflects that a part of adult males are not detected during count surveys. A null or negative recruitment rate would mean that immature males moulted before the survey or/and the assumption on the stability of the adult sex ratio is overrated.

Practically, the maximum recruitment rate is deduced by using its relationship to the maximum growth rate and the maximum survival, see Equation \@ref(eq:GR). This relationship comes from a simple reasoning for a closed population: the population size in year $t$ is equal to the number of adults that survived over year $t - 1$ plus the offspring produced in year $t - 1$ that survived until the reproduction period of year $t$, i.e. the recruitment in year $t$. The growth rate of a population is thus the sum of the adult survival rate and the recruitment rate [@Flint2015]. For an open population, adult survival and recruitment rate are confounded to adult and recruit migration respectively. This relationship becomes more complex if a species with delayed maturity is considered, see @Robertson2008.

@@

\begin{align} 
N_{t} & = N_{t-1} - D_{t-1} + R_{t} \notag \\
N_{t} & = N_{t-1} - (1 - s_{t}).N_{t-1} + r_{t}.N_{t-1} \notag \\
N_{t} & = s_{t}.N_{t-1} + r_{t}.N_{t-1} \notag \\
\frac{N_{t}}{N_{t-1}} & = s_{t} + r_{t} \notag \\
\lambda_{t} & = s_{t} + r_{t} \notag \\
r_{t} & = \lambda_{t} - s_{t}  
(\#eq:GR)
\end{align}

The maximum growth rate occurred for both populations during their expansion phase, when the evolution of the population size was the steepest. To estimate robust maximum population growth rates for both populations, we smoothed annual population growth rate over a consistent time period by using a linear regression on the logarithm scale, see Equation \@ref(eq:L). To do so, we rejected Great Britain data before 1972 since the size estimate of this population is noisy below 50 individuals (Figure \@ref(fig:count)). After reaching the threshold of 1000 individuals, the Great Britain population growth observed a strong inflexion whereas no harvest pressure was applied. This observation suggests that a negative density-dependence process might occur beyond 1000 individuals and led to consider only the first sequence of the time series to infer on the maximum growth rate in Great Britain, i.e. 1972-1981. For the French population, the sequence without harvest spread over the period 1994-1999.

@@

@@

Over the "high harvest" period, i.e. from 2006, control strategies differed from Great Britain to France. In Great Britain, harvest occurred mostly in winter, i.e. before the breeding period (53.4% of adults were shot before the 30^th May), whereas it was done mostly in summer in France, i.e. during and after the breeding period (81.6% of adults were shot after the 30^th May). We compared the population growth of the two populations in light of the difference between the two strategies and we assessed the respective response of the vital rates by comparing the average values over the "high harvest" period to proxies of maximum vital rates estimated when both populations reached their maximum growth (see Section \@ref(VAL)).

For all the sub-models in Section \@ref(MM), the Bayesian framework was used for its efficiency and simplicity to propagate error through the parameters. We used uninformative priors on all parameters. As the maximum growth rate is an life history trait quite stable among populations of a same species, we used a uninformative hierarchical prior for this parameter. We generated three chains of length 500000, with a thinning of 10 to avoid autocorrelation in the samples, and we discarded the first 2000 samples as burn-in. Chain convergence was assessed using the Gelman and Rubin convergence diagnostic (R<1.1, @Gelman1992). We fit the models using NIMBLE [@Valpine2017] run from R [@RCT2022]. Data and code are available here: https://github.com/adri-tab/Ruddy_duck_vital_rates. The values **X[Y; Z]** reported in results are the medians and the associated boundaries of the 95% confidence interval of posterior distributions. The median was preferred to the mean because of its robustness to skewed distribution.

@@

The harvest rate on adults is considered high for both populations from 2006 (Figure \@ref(fig:sup)). There is no significant difference between the two populations, t-test p-value : 0.23, and the average harvest rate is 0.49.

Time series derived from count data show that the two populations display comparable initial growth. They grew freely in both countries until 1999. Although shooting was initiated in 1999 in both countries, both populations begun declining only after much higher harvest efforts had been achieved in 2005 and 2018 respectively (figure 3). This history underlines that both populations observed a large spectrum of harvest effort, from no pressure to very high pressure. For sake of interpretation, we ascribed the time-period 2001-2004 to the no harvest category in France because the harvest pressure was negligible hand hence had no effect on the population growth.

Also, covering contrasted populations dynamics corresponding to different levels of harvesting pressure. In France, data before 2004 correspond to a period of steady population growth with (almost) no harvest pressure, whereas data in the following years correspond to a stabilized population size with a significant level of harvest pressure . In UK, the counting data correspond to a quick population depletion associated to a high level of harvest pressure, especially before the reproduction period. In parallel, the sampling dataset, which corresponds to individuals shot in winter during control operations, covers 9 years only in UK, with 5 years in common with the corresponding counting time series.
Estimating recruitment rate from the evolution of apparent sex-ratio in winter

The counting dataset, which differentiates the male-like from the female-like individuals in winter, covers 7 years and 19 years in UK and France respectively. Combined together, the counting datasets cover different population dynamics corresponding to a gradient of harvest pressure. In France, data before 2004 correspond to a period of population growth with (almost) no harvest pressure, whereas data in the following years correspond to a stabilized population size with a significant level of harvest pressure. In UK, the counting data correspond to a quick population depletion associated to a high level of harvest pressure, especially before the reproduction period. In parallel, the sampling dataset, which corresponds to individuals shot in winter during control operations, covers 9 years only in UK, with 5 years in common with the corresponding counting time series.

## Likelihood of the range of vital rates

########## OLD Recruitment rate variability

The male proportion in adults, which is necessary to infer on the recruitment rate from count data, is estimated at 0.60 [0.59; 0.61]. The counting method provides estimations of the recruitment rate, see Figure \@ref(fig:vital). In UK, values vary from 0.06 [0.05; 0.07] to 0.27 [0.20; 0.34] recruits per breeder, and in France, 0.08 [0.04; 0.12] to 0.76 [0.64; 0.88] recruits per breeder. There are significant year-to-year variations with a maximum amplitude of 0.68 recruits per breeder in France (9.5-time variability). On the available period, recruitment rate in UK is smaller than in France. Recruitment rates are estimated in both populations over 5 years, and no correlation is observed between the two populations. The lowest recruitment rate observed in France, which is in 2019, corresponds to the start of the depletion of the population. It is noticeable that it is in the same range than the UK values, also estimated during a strong population depletion.

Recruitment rates estimated from UK sample data are consistently higher than estimates from the corresponding count data, Figure \@ref(fig:vital). The correlation between the two methods is strong and points out that the sampling method overestimates the recruitment rate by 44% compared to the counting method, Figure \@ref(fig:vital). Estimation precision is similar for the two methods. The UK population started to deplete from 2006, but no significant shift is observed between the two periods.

```{r vital, fig.cap = "(ref:vital)"}
knitr::include_graphics("../Output/plot_paper_3.png")
```

## Likelihood of interannual variability in vital rates

########## Consistency of the corresponding adult survival

Adult survival rates were deduced from the recruitment rate estimates and the observed population growth rates, see Figure \@ref(fig:cor). In UK, values that are obtained from the counting method ranged from 0.21 [0.20; 0.22] to 0.68 [0.65; 0.72]. In France, survival rates vary from 0.39 [0.34; 0.44] to 1.08 [0.98; 1.17]. Like recruitment rates, there is significant year-to-year variability without clear temporal structure in both populations with a maximum amplitude of 0.69 in France (2.8-time variability). Unlike the recruitment rate, the survival rate in 2019 in France is not a low outlier.

As expected, survival rates estimated from UK sample data are consistently lower than estimates from the corresponding count data, Figure \@ref(fig:cor)A. The correlation between the two methods points out that the sampling method underestimates the survival rate by 32% compared to the counting method, Figure \@ref(fig:cor)B. Estimation precision is similar for the two methods. The sample time series shows that the adult survival is significantly lower after 2006, which corresponds to the start of the depletion of the UK population.

```{r cor, fig.cap = "(ref:cor)"}
knitr::include_graphics("../Output/plot_paper_4.png")
```

########### Comparison to maximum recruitment rate

Upon the conservative assumption of a maximum adult survival uniformly distributed over 0.70 to 0.90, the maximum recruitment rate reaches 0.65 [0.51; 0.75] and 0.66 [0.45; 0.90] recruits per breeder for UK and France respectively. When there is no harvest, the estimated average recruitment rate of the French population is in the range of the proxy of the maximum recruitment rate, 0.51 [0.46; 0.57] recruits per breeder. When there is harvest, the average recruitment rates are 0.14 [0.12; 0.16] and 0.33 [0.31; 0.36] recruits per breeder for the UK and France respectively. Both values are significantly lower than the maximum recruitment rate, and it is noticeable that the average recruitment rate is lower for the UK population, which is in depletion over the count time series, than for the French population, which of the dynamic is mostly stable over the monitored period.

## Response of vital rates to two different harvest strategies

```{r mgr, fig.cap = "(ref:mgr)"}
knitr::include_graphics("../Output/plot_paper_5_mod.png")
```

dire que disantangle recruitment rate to survival bring capital information to understand the process of harvest on the population dynamics and brings knowledge sur les leviers pour mieux conserver une espèce 
on mieux la controler for invasive species

dire que grâce à ça, on peut fournir des variations interannuelles de survie et de recruitment, ce que ne permet pas toujours la cmr en raison des tailles d'échantillon.

@@

@@

Assuming that the sample data provide a good picture of the interannual variability of the recruitment rate, the good correlation with the outputs of the counting method demonstrates the ability of this method to track the recruitment rate variability. This result is quite robust because the correlation is obvious with only a five-year time series. The assumption of a constant adult sex ratio over time is then acceptable because a significant change in adult sex ratio among years would have biased the evolution of the recruitment rate. Since this ratio integrates many age-cohorts, the temporal autocorrelation is structurally strong. Significant variations can then occur only every multiple years, especially if the species is long-living. Consequently, it is not necessary to monitor and update the adult sex ratio on a yearly base. The adult sex ratio found here on a restricted time period is in line with the ruddy duck in its native area [0.62 in @Bellrose1980] and with other duck species [@Wood2021]. The robustness of the correlation between the two methods might be overrated because two successive population size values are parts of the recruitment rate, which mechanically scale the outputs. It explains why the correlation on the immature proportion, the other component of the recruitment rate, is a bit more noisy than for the recruitment rate. The uncertainty around the estimates is similar for the two methods even though the counting method is more indirect to infer on the recruitment rate, which is satisfying.

@@

# Supplementary Materials

```{r sup, fig.cap = "(ref:sup)"}
knitr::include_graphics("../Output/plot_paper_6.png")
```

(ref:count) Evolution of the size of the two Ruddy duck populations; y-axis is on a log-scale; data availability refers to censuses that distinguished female-like from male-like individuals

@@

(ref:pfem) Female proportion in hunting bags vs in counts; For the first category, one point corresponds to the female proportion estimated from direct sex identification of the Ruddy ducks controlled over one year in a population; for the second category, one point is the proportion of female-like plumage in a population counted in winter; the difference is due to immature males that look like females before moulting

(ref:vital) Evolution of the size of the two Ruddy duck populations in light of the harvest pressure; on time periods without harvest, maximum population growth rates can be estimated (slopes in red for the UK and in blue for France)

(ref:cor) (A) Estimates of recruitment rate, i.e. average number of recruits produced per breeder; (B) Comparison of the recruitment rate values between the two estimation methods; bars define the 95% confidence intervals
